---
title: "Exercise_1"
output: html_document
date: "2023-03-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## STEP 2. Open the file in RStudio as a text file to clean up for import + import with `read_csv()`

Import the file using read_csv() function:

```{r Step 2, results = 'hide'}
library(readr)
my_connections <- read_csv("C:/Users/ulyan/OneDrive - McGill University/Documents/MMA/Winter II 2023/Org Network Analysis/Exercise 1/Basic_LinkedInDataExport_03-09-2023/Connections.csv")
```

## STEP 3. Get the count of your contacts by their current employer + total count

Import library:

```{r Step 3.1, results = FALSE}
library(tidyverse)
library(dplyr)
```

Count of contacts by current employer (excluding NA rows):

```{r Step 3.2}
contacts_by_employer <- drop_na(my_connections, "Company")
contacts_by_employer_count <- nrow(contacts_by_employer)

# Print the contacts by employer
print(contacts_by_employer_count)
```

Total count of contacts:

```{r Step 3.3}
total_count <- nrow(my_connections)

# Print the total count
print(total_count)

```


## STEP 4. Create nodes and edges dataframes to use with igraph

Install packages:

```{r Step 4.1}
library(tidygraph)
library(igraph)
```


# Pre-process data

Drop the "Email Address" column:

```{r Step 4.2}
my_connections <- my_connections %>% select(-"Email Address")
```


Drop any rows with missing values in the remaining columns:

```{r Step 4.3}
my_connections <- my_connections %>% drop_na()
```


# Create a tidygraph object from the connections data

```{r Step 4.4}
my_connections_graph <- my_connections %>%
  as_tbl_graph(nodes = c("First Name", "Last Name", "Company", "Position"),
               edges = c("First Name", "Last Name"),
               node_key = "name")
```



# Extract the nodes and edges dataframes from the tidygraph object

Note: 'nodes_df' first lists all given names and then all last names. This also causes issues with 'edges_df'. I couldn't figure out how to fix this. 

```{r Step 4.5}
nodes_df <- my_connections_graph %>%
  activate(nodes) %>%
  as_tibble()

edges_df <- my_connections_graph %>%
  activate(edges) %>%
  as_tibble()
```

Create a new edges dataframe with name values in place of indexes:

```{r Step 4.6}
edges_df2 <- edges_df %>%
  mutate(from = nodes_df$name[from],
         to = nodes_df$name[to])
```

## STEP 5. Plot the resulting network

Create an igraph object from the nodes and edges dataframes:

```{r Step 5.1}
my_igraph <- graph_from_data_frame(d = edges_df2, vertices = nodes_df)
```


Plot the resulting network:

```{r Step 5.2}
plot(my_igraph)
```
Note: Because of the issues with 'nodes_df' and 'edges_df', plot is displayed incorrectly and looks very busy.